---
title: "Getting Started with excelsioR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with excelsioR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

excelsioR helps you read and write Excel files, with special handling for messy,
real-world spreadsheets. It's built on top of tidyxl for reading (which preserves
cell-level information) and openxlsx for writing.

## Installation

```{r}
# Install from GitHub
devtools::install_github("ces0491/excelsioR")
```

```{r setup}
library(excelsioR)
library(dplyr)
```

## Writing Excel Files

### Basic Write Operations

Write a data frame to an Excel file:

```{r}
# Write mtcars to an Excel file
write_wb(
  mtcars,
  sheet_name = "cars",
  wb_dir = "my_data.xlsx",
 save_wb = TRUE
)
```

### Writing to Specific Cells

You can specify where to paste the data using Excel-style cell references.
excelsioR supports multi-letter columns (AA, AB, etc.):

```{r}
# Write starting at cell B5
write_wb(
  iris[1:10, ],
  sheet_name = "flowers",
  paste_coord = "B5",
  wb_dir = "positioned_data.xlsx",
  save_wb = TRUE
)

# Write to column AA (27th column)
write_wb(
  data.frame(x = 1:5),
  sheet_name = "Sheet1",
  paste_coord = "AA1",
  wb_dir = "wide_sheet.xlsx",
  save_wb = TRUE
)
```

### Multiple Sheets

Write multiple data frames to different sheets using a named list:

```{r}
# Create a list of data frames
my_sheets <- list(
  summary = data.frame(metric = c("Total", "Average"), value = c(100, 50)),
  details = mtcars[1:10, ],
  metadata = data.frame(created = Sys.Date(), author = "Me")
)

# Write all sheets at once
write_wb(my_sheets, wb_dir = "multi_sheet_report.xlsx", save_wb = TRUE)
```

### Time Series Data

zoo and xts objects are automatically converted with a date column:

```{r}
library(zoo)

# Create a zoo time series
dates <- seq(as.Date("2024-01-01"), by = "day", length.out = 30)
values <- data.frame(
  price = cumsum(rnorm(30)),
  volume = rpois(30, 100)
)
ts_data <- zoo(values, dates)

# Write to Excel - date column is added automatically
write_wb(
  ts_data,
  sheet_name = "daily_data",
  wb_dir = "timeseries.xlsx",
  save_wb = TRUE
)
```

## Reading Excel Files

### Basic Read Operations

Read all Excel files from a directory:

```{r}
# Read all Excel files
data <- read_wb("path/to/excel/files")

# View the structure
data
#> # A tibble: 3 x 3
#>   file_name    raw_excel_data all_excel_data_tbl
#>   <chr>        <list>         <list>
#> 1 sales_2024   <named list>   <tibble>
#> 2 inventory    <named list>   <tibble>
#> 3 forecast     <named list>   <tibble>
```

### Filtering Files and Sheets

```{r}
# Read specific workbooks
data <- read_wb(
  source_data_dir = "path/to/files",
  reqd_wkbks = c("sales_2024", "inventory")
)

# Read specific sheets from all workbooks
data <- read_wb(
  source_data_dir = "path/to/files",
  reqd_sheets = c("Summary", "Q4")
)
```

### Working with the Output

The output has three columns:

- `file_name`: Cleaned file name (no extension, special characters replaced)
- `raw_excel_data`: Nested list of raw tidyxl cell data per sheet
- `all_excel_data_tbl`: Nested tibble of processed tidy data

```{r}
# Extract processed data for one file
sales_data <- data %>%
  filter(file_name == "sales_2024") %>%
  pull(all_excel_data_tbl) %>%
  .[[1]]

# The tidy data has columns: sheet, row, col, numeric, character, date
head(sales_data)
#>   sheet row col  numeric character       date
#> 1 Sheet1   2   2   1000.5    Revenue 2024-01-01
#> 2 Sheet1   3   2   2500.0    Revenue 2024-02-01
```

### Safe Reading

By default, `read_wb()` copies files to a temp directory before reading. This
prevents accidental corruption of your source files:

```{r}
# Files are copied to a timestamped temp folder
data <- read_wb("important/data")
#> No dest_data_dir specified. Source data will be copied to:
#>   C:/Users/.../Temp/read_wb_data/source_data_copy_20240115
#> This copied data will be cleared when your session restarts.

# Or specify your own destination
data <- read_wb(
  "important/data",
  dest_data_dir = "working/copies"
)
```

## Password-Protected Files

### Reading Protected Files

```{r}
# Will prompt for password in RStudio
data <- read_wb("protected/files", password_protected = TRUE)
```

### Unlocking Files Directly

```{r}
# Prepare file info
files_to_unlock <- data.frame(
  file_path = c("file1.xlsx", "file2.xlsx"),
  file_name = c("file1", "file2")
)

# Unlock (will prompt for password)
unlock_wb(files_to_unlock)

# Or provide password directly (less secure)
unlock_wb(files_to_unlock, wb_password = "secret123")
```

### Backend Selection

excelsioR supports two backends for password protection:

```{r}
# Auto-select (prefers rpxl if available)
unlock_wb(files, wb_password = "secret")

# Force rpxl (Python-based, no Java)
unlock_wb(files, wb_password = "secret", backend = "rpxl")

# Force XLConnect (Java-based)
unlock_wb(files, wb_password = "secret", backend = "xlconnect")
```

Install backends:

```{r}
# rpxl (recommended - lighter weight)
remotes::install_github("epicentre-msf/rpxl")

# XLConnect (requires Java)
install.packages("XLConnect")
```

## Tips and Best Practices

### 1. Use the Raw Data for Complex Spreadsheets

The `all_excel_data_tbl` column makes assumptions about your data structure.
For complex or unusual layouts, work with `raw_excel_data` directly:

```{r}
# Get raw tidyxl data
raw <- data$raw_excel_data[[1]]$Sheet1

# This has all cell information: row, col, data_type, formula, etc.
raw %>%
  filter(!is_blank) %>%
  select(row, col, character, numeric, date)
```

### 2. Batch Processing

Process multiple files efficiently:

```{r}
# Read all files
all_data <- read_wb("monthly_reports")

# Process each file's data
results <- all_data %>%
  mutate(
    summary = purrr::map(all_excel_data_tbl, ~ {
      .x %>%
        group_by(character) %>%
        summarise(total = sum(numeric, na.rm = TRUE))
    })
  )
```

### 3. Writing Reports

Build up a workbook incrementally:

```{r}
# Create workbook
wb <- write_wb(summary_data, sheet_name = "Summary", wb_dir = "report.xlsx")

# Add more sheets to the same workbook
wb <- write_wb(detail_data, sheet_name = "Details", wb = wb)
wb <- write_wb(chart_data, sheet_name = "Charts", wb = wb)

# Save at the end
openxlsx::saveWorkbook(wb, "report.xlsx", overwrite = TRUE)
```
