---
title: "Taming Messy Spreadsheets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taming Messy Spreadsheets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
  comment = "#>"
)
library(excelsioR)
library(dplyr)
library(tidyr)
```

## Two Kinds of Spreadsheets

There's a fundamental difference between:

1. **Data spreadsheets** - Tables designed to store data: headers in row 1,
   one record per row, consistent columns. readxl handles these brilliantly.

2. **Presentation spreadsheets** - Documents designed for humans: titles,
   logos, merged cells, footnotes, multiple tables per sheet. These look
   great printed but require more care to parse.

Many spreadsheets in the wild are presentation spreadsheets:

```text
     A          B          C          D          E
1   [Company Logo]        Q3 2024 Sales Report
2
3                    Jan        Feb        Mar
4   North         $12,450    $13,200    $14,100
5   South          $9,800    $10,100    $11,200
6   East          $11,300    $11,800    $12,400
7   West          $10,200    $10,500    $10,900
8
9   Notes: Figures in USD. Excludes returns.
```

With readxl, you'd specify `skip = 2`, `n_max = 4`, and clean up column names.
That works well for one file - but what about 50 files with varying layouts?

## Building on tidyxl

[tidyxl](https://nacnudus.github.io/tidyxl/) is an excellent package that reads
every cell with its exact position, formatting, and value. This gives you
complete control over extraction from any spreadsheet structure.

excelsioR builds on tidyxl to provide a complete workflow:

- **Batch processing**: Read entire folders of files at once
- **Structured output**: Nested tibbles you can unnest and filter
- **Write capabilities**: Send results back to Excel with openxlsx
- **Password support**: Unlock protected workbooks automatically

### Creating a Reproducible Example

Let's create a messy spreadsheet to demonstrate:

```{r create-messy-data}
library(excelsioR)
library(openxlsx)

# Create a temp directory for our examples
demo_dir <- tempdir()

# Build a "human-friendly" but analysis-hostile spreadsheet
wb <- createWorkbook()
addWorksheet(wb, "Sales Report")

# Row 1: Title (not headers!)
writeData(wb, "Sales Report", x = "Q3 2024 Regional Sales", startRow = 1, startCol = 2)

# Row 3: Actual headers
writeData(wb, "Sales Report", x = data.frame(
  Region = character(), Jan = numeric(), Feb = numeric(), Mar = numeric()
), startRow = 3)

# Row 4-7: Data
sales_data <- data.frame(
  Region = c("North", "South", "East", "West"),
  Jan = c(12450, 9800, 11300, 10200),
  Feb = c(13200, 10100, 11800, 10500),
  Mar = c(14100, 11200, 12400, 10900)
)
writeData(wb, "Sales Report", x = sales_data, startRow = 4, colNames = FALSE)

# Row 9: Footnote
writeData(wb, "Sales Report", x = "Notes: Figures in USD", startRow = 9, startCol = 1)

# Save it
messy_file <- file.path(demo_dir, "messy_sales.xlsx")
saveWorkbook(wb, messy_file, overwrite = TRUE)

cat("Created:", messy_file, "\n")
```

### Default Behavior

```{r standard-read, error=TRUE}
# Default read without specifying structure:
readxl::read_excel(messy_file)
```

Without hints about the structure, the first row becomes column names. You
could specify `skip = 2`, `n_max = 4` to extract the data correctly. For a
single file, that works great.

But with many files having different layouts, cell-level access is useful:

### Cell-Level Access with excelsioR

```{r excelsior-read}
# Read using excelsioR's tidyxl-based approach
result <- read_wb(
  source_data_dir = demo_dir,
  reqd_wkbks = "messy_sales"
)

# What we get back
result
```
The `raw_excel_data` column contains every cell with its exact position.
The `all_excel_data_tbl` column contains processed data based on common
spreadsheet patterns.

```{r examine-raw}
# Look at the raw cell data (first file, first sheet)
raw_cells <- result$raw_excel_data[[1]][[1]]
head(raw_cells, 10)
```

Now you have complete control. You can:

1. Filter to only the data rows (4-7)
2. Ignore the title and footnote
3. Reconstruct the table exactly as intended

```{r extract-clean}
# Extract just the data we want
clean_data <- raw_cells %>%
  filter(row >= 4, row <= 7, !is_blank) %>%
  select(row, col, numeric, character) %>%
  mutate(value = coalesce(as.character(numeric), character)) %>%
  select(row, col, value) %>%
  pivot_wider(names_from = col, values_from = value)

# Rename columns properly
names(clean_data) <- c("row_num", "Region", "Jan", "Feb", "Mar")
clean_data <- clean_data %>%
  select(-row_num) %>%
  mutate(across(Jan:Mar, as.numeric))

clean_data
```

## Batch Processing: The Real Power

The true value of excelsioR emerges when you have many files:

```{r batch-demo}
# Create multiple messy files with different structures
batch_dir <- file.path(tempdir(), "batch_demo")
dir.create(batch_dir, showWarnings = FALSE)
regions <- c("north", "south", "east", "west")

for (region in regions) {
  wb <- createWorkbook()
  addWorksheet(wb, "Data")

  # Each file has slightly different layout
  start_row <- 2

  writeData(wb, "Data", x = paste(toupper(region), "Region Report"),
            startRow = 1, startCol = 1)

  monthly_data <- data.frame(
    Month = month.abb[1:3],
    Sales = c(10000, 11000, 12000),
    Returns = c(200, 250, 300)
  )

  writeData(wb, "Data", x = monthly_data, startRow = start_row)
  saveWorkbook(wb, file.path(batch_dir, paste0(region, "_report.xlsx")),
               overwrite = TRUE)
}

# One function call reads all files
all_reports <- read_wb(batch_dir)
all_reports

# See what files we got
all_reports$file_name
```

Compare this to the manual approach:

```r
# Without excelsioR (don't run this)
files <- list.files("path/to/files", pattern = "\\.xlsx$", full.names = TRUE)
all_data <- list()
for (f in files) {
  sheets <- readxl::excel_sheets(f)
  for (s in sheets) {
    # But wait - which row has headers? Different for each file...
    # And some files are password protected...
    # And some have merged cells...
    all_data[[length(all_data) + 1]] <- readxl::read_excel(f, sheet = s)
  }
}
# Now somehow combine these misaligned data frames...
```

## Writing Results Back

After analysis, write results to Excel for stakeholders who prefer spreadsheets:

```{r write-demo}
# Summarize the data
summary_data <- data.frame(
  Region = c("North", "South", "East", "West"),
  Q1_Total = c(39750, 31100, 35500, 31600),
  Avg_Monthly = c(13250, 10367, 11833, 10533)
)

# Write single sheet
write_wb(
  r_data = summary_data,
  sheet_name = "Summary",
  wb_dir = file.path(demo_dir, "analysis_output.xlsx"),
  save_wb = TRUE
)

# Write multiple sheets from a named list
multi_sheet_data <- list(
  "Executive Summary" = summary_data,
  "North Detail" = data.frame(Month = month.abb[1:3], Sales = c(12450, 13200, 14100)),
  "South Detail" = data.frame(Month = month.abb[1:3], Sales = c(9800, 10100, 11200))
)

write_wb(
  r_data = multi_sheet_data,
  wb_dir = file.path(demo_dir, "full_report.xlsx"),
  save_wb = TRUE
)
```

### Writing to Specific Locations

Need to paste data into an existing template? Use `paste_coord`:

```{r paste-coord}
# Write to cell AA10 (multi-letter columns supported)
write_wb(
  r_data = data.frame(x = 1:3, y = 4:6),
  sheet_name = "Template",
  paste_coord = "C5",  # Start at column C, row 5
  wb_dir = file.path(demo_dir, "template_output.xlsx"),
  save_wb = TRUE
)
```

## Common Messy Spreadsheet Patterns

excelsioR helps with these common issues:

| Pattern | Problem | Solution |
| ------- | ------- | -------- |
| Title rows | Headers aren't in row 1 | Use `raw_excel_data` to find actual headers |
| Footnotes | Extra rows after data | Filter by row position |
| Merged cells | Values only in top-left | tidyxl tracks merge ranges |
| Multiple tables | Several datasets per sheet | Filter by row/col ranges |
| Mixed types | Numbers stored as text | Access `numeric`, `character`, `date` columns separately |
| Inconsistent layouts | Each file structured differently | Programmatic extraction per file |

## Key Takeaways

1. **Human-friendly != analysis-friendly**: Spreadsheets designed for reading
   are often hard to parse programmatically

2. **Cell-level access matters**: When structure varies, you need the raw cells,
   not a guessed data frame

3. **Batch processing saves time**: Reading one file is easy; reading 50 files
   with different layouts is where excelsioR shines

4. **Keep Excel as the interface**: Your stakeholders get the spreadsheets they
   want; you get reproducible R code

## Next Steps

- See `?read_wb` for all reading options
- See `?write_wb` for writing options including time series support
- See `?unlock_wb` for password-protected files
